---
title: 'Do environmental interventions ameliorate gut dysbiosis in Huntington disease mice?'
subtitle: "Data Analysis"
author: 'Saritha Kodikara'
#date: 'November 9, 2017'
date: '\today'
#date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    toc: true   # can remove
    toc_depth: 3 # can remove
    number_sections: true
  highlight: zenburn
header-includes:
- \usepackage{fancyhdr}
- \usepackage{mathtools}
- \usepackage{xcolor, hyperref}
- \usepackage{lipsum}
- \usepackage{subfig}
- \usepackage{float}
- \setlength{\headheight}{34pt}
- \setlength{\footskip}{25pt}
- \pagestyle{fancy}
- \renewcommand{\headrulewidth}{0.5pt}
- \renewcommand{\footrulewidth}{0.5pt}
- \lhead{\includegraphics[width=8cm,height=1cm]{logo-unimelb}}
- \cfoot{\scriptsize School of Mathematics and Statistics, Melbourne Integrative Genomics \\ The University of Melbourne}
- \rhead{\thepage}
- \hypersetup{colorlinks   = true, linkcolor=blue, urlcolor  = blue}
- \fancypagestyle{plain}{\pagestyle{fancy}}
editor_options: 
  chunk_output_type: console
bibliography: Ref.bib
---

```{r global_options, include=FALSE}
library(knitr)
# global options to show by default the code, dump the figures into /Figures etc
knitr::opts_chunk$set(dpi = 100, 
                      echo=TRUE, 
                      warning=FALSE, message=FALSE, eval = TRUE,
                      fig.show=TRUE, fig.width= 5,fig.height= 5,fig.align='center', out.width = '50%', fig.path= 'Figures/')
library(dplyr)
library(readxl)
library(magrittr)
library(ggplot2)
library(ggpubr)
library(car)
library(nlme)
library(stargazer)
library(dotwhisker)
library(sjPlot)
require(xtable)
library(TSA)
library(jtools)
library(emmeans)
library(graphics)
library(tidyr)
library(ordinal)
library(viridis)
library(lmerTest)
library(ggpubr)
library(rstatix)
library(patchwork)
library(kableExtra)
```


<!-- the {-} indicate that you do not wish to number this section -->
# Executive Summary {-}
- The data were preprocessed to remove any samples with more than 80% missing data which resulted in removing four mice from the study.

- Except for the clinical variable clasping score, all other single time-point observations, such as Colon length, were modelled using Linear mixed models (LMMs). A LMM is an extension of a linear model to incorporate both fixed-effect terms and random effect terms. For repeated measures, time was added as covariate in the LMM. All two-way interactions between fixed effects were included in the LMM models.

- For the variable clasping score measured on an ordinal scale, we fitted a cumulative linear mixed model with Laplace approximation.

- For models with at least one significant effect, we performed a post-hoc contrast analysis on estimated marginal means to identify significant pairs using the emmeans R package [@emmeans]. The emmeans method performs pairwise comparisons on the levels of significant main effects while controlling for other factors. For example, if genotype was significant in the LMM, a pairwise comparisons between two genotypes separately for each sex and housing type would be conducted. All the p-values from the pairwise comparisons were then adjusted for multiple comparisons using Tukey adjustment. 



# Data preprocessing (Removing insufficient samples)

Data from '210330_Gubert_EE-EX_Data_All.xlsx' were analysed by combining the male and female details from  'Female_EEEX_Data_Final' and 'Male_EEEX_Data_Final'.

First, the samples with more than 80\% missing data}were removed from the analysis (insufficient data). As a result, 4 mice were removed from the analysis. These four samples did not have any clinical information ( i.e. Colon Length, Gut Permeability, Fecal Water content at Week 12, Isobutyrate, 2-Methyl butyrate, Isovalerate). Those mice fell into the following combination of Sex, genotype and housing condition:\
-- Female/ HD/ EE \qquad
-- Female/ HD/ EX \qquad
-- Female/ WT/ SH \qquad
-- Male/ HD/ EX \qquad

```{r, echo=FALSE}
data_Female<-read_excel("data/210330_Gubert_EE-EX_Data_All.xlsx","Female_EEEX_Data_Final")
#converting water content into the same format as Males
data_Female$FH20_Wk7<-data_Female$FH20_Wk7*100
data_Female$FH20_Wk8<-data_Female$FH20_Wk8*100
data_Female$FH20_Wk9<-data_Female$FH20_Wk9*100
data_Female$FH20_Wk10<-data_Female$FH20_Wk10*100
data_Female$FH20_Wk11<-data_Female$FH20_Wk11*100
data_Female$FH20_Wk12<-data_Female$FH20_Wk12*100
#Counting the number of columns with missing values
countNA <- function(df) apply(df, MARGIN = 1, FUN = function(x) length(x[is.na(x)]))
#Filter out sample with more than 80% missing values
new_f_data<-data_Female %>% filter(countNA(data_Female)<60)
# Extracting Colon length columns for the analysis
data_female_clean<-new_f_data[,2:4] %>%mutate(Gender=rep("Female",69)) %>%
  mutate(new_f_data[,6],new_f_data[,c(34,35,42:45,48,49,54,55,58:65,68:75)])

data_Male<-read_excel("data/210330_Gubert_EE-EX_Data_All.xlsx","Male_EEEX_Data_Final")
#converting water content into nearest integer
data_Male$FH20_Wk12<-round(data_Male$FH20_Wk12)
#Counting the number of columns with missing values
countNA <- function(df) apply(df, MARGIN = 1, FUN = function(x) length(x[is.na(x)]))
#Filter out sample with more than 80% missing values
new_m_data<-data_Male %>% filter(countNA(data_Male)<60)
# Extracting interested columns for the analysis
data_male_clean<-new_m_data[,c(2,4:5)] %>%mutate(Gender=rep("Male",70)) %>%
  mutate(new_m_data[,1],new_m_data[,c(65, 59,64,60,61,63,39,40,45,46,49:56,66:73)])
#colnames(data_female_clean);colnames(data_male_clean)
colnames(data_female_clean)<-colnames(data_male_clean)

# Combine both Males and Females
data_clean<-rbind(data_female_clean, data_male_clean)

#Converting Genotype, Housing, Gender and Box to be factor variables
# When creating the factors put the reference category first in the levels argument 
data_clean$Genotype<-factor(data_clean$Genotype,levels = c("WT", "HD"))
data_clean$Housing <-factor(data_clean$Housing , levels = c("SH","EE","EX"))
data_clean$Gender<-factor(data_clean$Gender, levels = c("Female","Male"))
data_clean$Box <-factor(data_clean$Box )
colnames(data_clean)[which(names(data_clean) == "Gender")] <- "Sex"
```

After the preprocessing, the composition of the samples are as follows:

```{r echo=FALSE}
count_table <- xtabs(~Genotype+Housing+Sex, data=data_clean)
ftable(count_table) # print table
```



# Effect of Housing, Genotype and Sex on Brain Weight

In this section, we use LMM to find the significant variables that are associated with the Brain weight of a mouse. LMM model considers both fixed and random effects. Fixed effects are variables that we expect will have an impact on our response variable. For example, we expect Sex, genotype and housing would impact the brain weight of a mouse. Thus, those variables (i.e. Sex, genotype and housing) are considered fixed effects. Random effects are usually grouping factors that may have an impact on the response variable but we are not specifically interested in their impact. In this study, mice are housed within cages (random effect), while we are investigating the effect of sex (male, female); genotype (WT, HD) and housing (SH, EE, EX) (fixed effects). 
```{r, echo=FALSE}
m1 <- lme(Brain_Weight ~ Genotype + Housing+ 
                    Sex+Genotype:Sex+Housing:Sex+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Brain_Weight)))

#summary(mixed.lmer)

#tab_model(mixed.lmer)

stargazer(m1, type = "text",
          report = ('vc*p'),
          digits = 3,
          star.char = c("+", "*", "**", "***"),
          star.cutoffs = c(.1, .05, .01, .001),
          digit.separator = "", notes = c("+ p<0.1; * p<0.05; ** p<0.01; *** p<0.001", "Constants are significant (p<0.001)"), notes.append = FALSE,dep.var.labels="Brain weight")
```

- Based on the LMM, significant variables are:\
\qquad - Sex
\qquad - Genotype

- A significant parameter with a positive value (i.e $>0$) indicates a positive effect on the average brain weight. For example, the average brain weight is higher for male mice than female mice.




# LMM for all interested variables

```{r, echo=FALSE}
M1 <- lme(Gut_Permeability_FITC ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Gut_Permeability_FITC)))
M2 <- lme(Gut_Transit_Time ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Gut_Transit_Time)))
M3 <- lme(Colon_Length~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Colon_Length)))
M4 <- lme(Caecum_Weight ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Caecum_Weight)))

stargazer(M1,M2,M3,M4, type = "text",
          report = ('vc*p'),
          digits = 3,
          star.char = c("+", "*", "**", "***"),
          star.cutoffs = c(.1, .05, .01, .001),
          digit.separator = "", notes = c("+ p<0.1; * p<0.05; ** p<0.01; *** p<0.001", "Constants are significant (p<0.001)"), notes.append = FALSE,
         dep.var.labels=c("Gut permeability", "Gut transit time","Colon length","Caecum weight"))
```
\newpage
```{r, echo=FALSE}

M1 <- lme(Caecum_Length ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Caecum_Length)))
M2 <- lme(DG_Swing_Fore ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Swing_Fore)))
M3 <- lme(DG_Swing_Hind ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Swing_Hind)))
M4 <- lme(DG_Stride_Fore ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Stride_Fore)))

stargazer(M1,M2,M3,M4, type = "text",
          report = ('vc*p'),
          digits = 3,
          star.char = c("+", "*", "**", "***"),
          star.cutoffs = c(.1, .05, .01, .001),
          digit.separator = "", notes = c("+ p<0.1; * p<0.05; ** p<0.01; *** p<0.001", "Constants are significant (p<0.001)"), notes.append = FALSE,
         dep.var.labels=c("Caecum length", "Swing time forepaw","Swing time hind paw","Stride time forepaw"))

```
\newpage
```{r, echo=FALSE}

M1 <- lme(DG_Stride_Hind ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Stride_Hind)))
M2 <- lme(DG_Stride_Length_Fore ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Stride_Length_Fore)))
M3 <- lme(DG_Stride_Length_Hind ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Stride_Length_Hind)))

stargazer(M1,M2,M3, type = "text",
          report = ('vc*p'),
          digits = 3,
          star.char = c("+", "*", "**", "***"),
          star.cutoffs = c(.1, .05, .01, .001),
          digit.separator = "", notes = c("+ p<0.1; * p<0.05; ** p<0.01; *** p<0.001", "Constants are significant (p<0.001)"), notes.append = FALSE,
         dep.var.labels=c("Stride time hind paw", "Stride length forepaw","Stride length hindpaw"))

```
\newpage
```{r, echo=FALSE}

M1 <- lme(DG_Absolute_Paw_Angle_Fore ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Absolute_Paw_Angle_Fore)))
M2 <- lme(DG_Absolute_Paw_Angle_Hind ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Absolute_Paw_Angle_Hind)))
stargazer(M1,M2, type = "text",
          report = ('vc*p'),
          digits = 3,
          star.char = c("+", "*", "**", "***"),
          star.cutoffs = c(.1, .05, .01, .001),
          digit.separator = "", notes = c("+ p<0.1; * p<0.05; ** p<0.01; *** p<0.001", "Constants are significant (p<0.001)"), notes.append = FALSE,
         dep.var.labels=c("Absolute paw angle forepaw", "Absolute paw angle hindpaw"))
```
\newpage
```{r, echo=FALSE}
M1 <- lme(DG_Stance_Width_Fore ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Stance_Width_Fore)))
M2 <- lme(DG_Stance_Width_Hind ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Stance_Width_Hind)))


stargazer(M1,M2,  type = "text",
          report = ('vc*p'),
          digits = 3,
          star.char = c("+", "*", "**", "***"),
          star.cutoffs = c(.1, .05, .01, .001),
          digit.separator = "", notes = c("+ p<0.1; * p<0.05; ** p<0.01; *** p<0.001", "Constants are significant (p<0.001)"), notes.append = FALSE, 
         dep.var.labels=c("Stance width forepaw", "Stance width hindpaw"))

```

\newpage
```{r, echo=FALSE}

M1 <- lme(DG_Propel_Brake_Ratio_FORE ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Propel_Brake_Ratio_FORE)))
M2 <- lme(DG_Propel_Brake_Ratio_HIND ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Propel_Brake_Ratio_HIND)))

stargazer(M1,M2,  type = "text",
          report = ('vc*p'),
          digits = 3,
          star.char = c("+", "*", "**", "***"),
          star.cutoffs = c(.1, .05, .01, .001),
          digit.separator = "", notes = c("+ p<0.1; * p<0.05; ** p<0.01; *** p<0.001", "Constants are significant (p<0.001)"), notes.append = FALSE, 
         dep.var.labels=c("propel:brake ratio forepaw","propel:brake ratio hind paw"))

```
\newpage
```{r, echo=FALSE}

M1 <- lme(Acetate ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Acetate)))

M2 <- lme(Proprionate ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Proprionate)))

M3 <- lme(Isobutyrate ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Isobutyrate)))
M4 <- lme(Butyrate ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Butyrate)))

stargazer(M1,M2,M3,M4, type = "text",
          report = ('vc*p'),
          digits = 3,
          star.char = c("+", "*", "**", "***"),
          star.cutoffs = c(.1, .05, .01, .001),
          digit.separator = "", notes = c("+ p<0.1; * p<0.05; ** p<0.01; *** p<0.001", "Constants are significant (p<0.001)"), notes.append = FALSE)

```
\newpage


```{r, echo=FALSE}

M1<- lme(Methylbutyrate2 ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Methylbutyrate2)))


M2 <- lme(Isovalerate ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Isovalerate)))
M3 <- lme(Valerate ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Valerate)))
M4 <- lme(Caproate ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Caproate)))

stargazer(M1,M2,M3,M4, type = "text",
          report = ('vc*p'),
          digits = 3,
          star.char = c("+", "*", "**", "***"),
          star.cutoffs = c(.1, .05, .01, .001),
          digit.separator = "", notes = c("+ p<0.1; * p<0.05; ** p<0.01; *** p<0.001", "Constants are significant (p<0.001)"), notes.append = FALSE)

```
\newpage

# LMM with Week as a covariate

```{r, echo=FALSE}
data_female_clean_time<-new_f_data[,2:4] %>%mutate(Gender=rep("Female",69)) %>%
  mutate(new_f_data[,c(6:20,28:33,36:41)])
data_male_clean_time<-new_m_data[,c(2,4:5)] %>%mutate(Gender=rep("Male",70)) %>%
  mutate(new_m_data[,1],new_m_data[,c(13:38)])
#colnames(data_female_clean);colnames(data_male_clean)
colnames(data_male_clean_time)<-colnames(data_female_clean_time)

# Combine both Males and Females
data_clean_time<-rbind(data_female_clean_time, data_male_clean_time)

#Converting Genotype, Housing, Gender and Box to be factor variables
data_clean_time$Genotype<-factor(data_clean_time$Genotype, levels=c("WT","HD"))
data_clean_time$Housing <-factor(data_clean_time$Housing ,  levels = c("SH","EE","EX"))
data_clean_time$Gender<-factor(data_clean_time$Gender, levels=c("Female", "Male"))
data_clean_time$Box <-factor(data_clean_time$Box )

colnames(data_clean_time)[which(names(data_clean_time) == "Gender")] <- "Sex"

long_data<-data_clean_time %>% 
  gather(v, value, Weight_Wk6:FH20_Wk12) %>% 
  separate(v, c("col", "Week"),sep="_Wk") %>% 
  arrange(MOUSE_ID) %>% 
  spread(col, value)


long_data$Week <-as.numeric(long_data$Week)
long_data_sorted<-long_data[order(long_data$MOUSE_ID,long_data$Box,long_data$Week),]
#Filtering out Week 6 from Rotarod data
rotarod_data<-long_data_sorted[,c(1:6,9)]%>%filter(Week!=6)

m1 <- lme(Weight ~ Genotype + Housing+ Sex+Week+
            Genotype:Sex+Sex:Housing+
            Genotype:Week+ Sex:Week + Housing:Week+Housing:Genotype,
          random =(~1|Box), data = subset(long_data_sorted, !is.na(Weight)))
m2 <- lme(Rotarod ~ Genotype + Housing+ Sex+Week+
            Genotype:Sex+Sex:Housing+
            Genotype:Week+ Sex:Week + Housing:Week+Housing:Genotype,
          random =(~1|Box), data = subset(rotarod_data, !is.na(Rotarod)))
m3<- lme(FH20 ~ Genotype + Housing+ Sex+Week+
            Genotype:Sex+Sex:Housing+
            Genotype:Week+ Sex:Week + Housing:Week+Housing:Genotype,
          random =(~1|Box), data = subset(long_data_sorted, !is.na(FH20)))
m4 <- lme(FOutput ~ Genotype + Housing+ Sex+Week+
            Genotype:Sex+Sex:Housing+
            Genotype:Week+ Sex:Week + Housing:Week+Housing:Genotype,
          random =(~1|Box), data = subset(long_data_sorted, !is.na(FOutput)))

######################## Food and Water intake per box by Gender#######

data_intake_Male<-read_excel("data/210330_Gubert_EE-EX_Data_All.xlsx","Male_EEEX_Data_Final_Food_H20")
#Sorting the data by Box number and inclusding a Sex column
data_intake_Male_sorted<-data_intake_Male[order(data_intake_Male$Box),-4]
data_intake_Male_sorted<-data_intake_Male_sorted%>%mutate(Sex=rep("Male", length(data_intake_Male_sorted$Box)), .after=Housing)
  
data_intake_Female<-read_excel("data/210330_Gubert_EE-EX_Data_All.xlsx","Female_EEEX_Data_Final_Food_H20",skip=1)


# Matching the Column order to Males and inclusding a Sex column
data_intake_Female_sorted<-data_intake_Female[,c(3,1,2,4:15)]
data_intake_Female_sorted<-data_intake_Female_sorted%>%mutate(Sex=rep("Female", length(data_intake_Female_sorted$...3)), .after=...2)

#colnames(data_intake_Female_sorted);colnames(data_intake_Male_sorted)
colnames(data_intake_Female_sorted)<-colnames(data_intake_Male_sorted)

food_intake<-rbind(data_intake_Female_sorted,data_intake_Male_sorted)

long_data_food<-food_intake %>% 
  gather(v, value, Food_Intake_Wk6:Water_Intake_Wk11) %>% 
  separate(v, c("col", "Week"), sep="_Wk") %>% 
  arrange(Box) %>% 
  spread(col, value)

long_data_food$Week <-as.numeric(long_data_food$Week)
food_sorted<-long_data_food[order(long_data_food$Box,long_data_food$Week),]

food_sorted$Genotype<-factor(food_sorted$Genotype, levels=c("WT","HD"))
food_sorted$Housing <-factor(food_sorted$Housing ,  levels = c("SH","EE","EX"))
food_sorted$Sex<-factor(food_sorted$Sex, levels=c("Female", "Male"))
food_sorted$Box <-factor(food_sorted$Box )

m5<- lme(Food_Intake ~ Genotype + Housing+ Sex+Week+
            Genotype:Sex+Sex:Housing+
            Genotype:Week+ Sex:Week + Housing:Week+Housing:Genotype,
          random =(~1|Box), data = subset(food_sorted, !is.na(Food_Intake)))
m6 <- lme(Water_Intake ~ Genotype + Housing+ Sex+Week+
            Genotype:Sex+Sex:Housing+
            Genotype:Week+ Sex:Week + Housing:Week+Housing:Genotype,
          random =(~1|Box), data = subset(food_sorted, !is.na(Water_Intake)))



stargazer(m5,m6,m1,m2,m3,m4, type = "text",
          report = ('vc*p'),
          digits = 3,
          star.char = c("+", "*", "**", "***"),
          star.cutoffs = c(.1, .05, .01, .001),
          digit.separator = "", notes = c("+ p<0.1; * p<0.05; ** p<0.01; *** p<0.001", "Constants are significant (p<0.001)"), notes.append = FALSE)


```

# Modelling the variable clasping

```{r, echo=FALSE}
data_female_clasping<-new_f_data[,2:4] %>%mutate(Gender=rep("Female",69)) %>%
  mutate(new_f_data[,6],new_f_data[,c(21:27)])
data_male_clasping<-new_m_data[,c(2,4:5)] %>%mutate(Gender=rep("Male",70)) %>%
  mutate(new_m_data[,1],new_m_data[,c(6:12)])
#colnames(data_female_clean);colnames(data_male_clean)
colnames(data_male_clasping)<-colnames(data_female_clasping)

# Combine both Males and Females
data_clasping<-rbind(data_female_clasping, data_male_clasping)

#Converting Genotype, Housing, Gender and Box to be factor variables
data_clasping$Genotype<-factor(data_clasping$Genotype, levels=c("WT","HD"))
data_clasping$Housing <-factor(data_clasping$Housing ,  levels = c("SH","EE","EX"))
data_clasping$Gender<-factor(data_clasping$Gender, levels=c("Female", "Male"))
data_clasping$Box <-factor(data_clasping$Box )
data_clasping$Clasping_Wk6<-factor(data_clasping$Clasping_Wk6, ordered = TRUE , levels = c(0:4))
data_clasping$Clasping_Wk7<-factor(data_clasping$Clasping_Wk7, ordered = TRUE , levels = c(0:4))
data_clasping$Clasping_Wk8<-factor(data_clasping$Clasping_Wk8, ordered = TRUE , levels = c(0:4))
data_clasping$Clasping_Wk9<-factor(data_clasping$Clasping_Wk9, ordered = TRUE , levels = c(0:4))
data_clasping$Clasping_Wk10<-factor(data_clasping$Clasping_Wk10, ordered = TRUE , levels = c(0:4))
data_clasping$Clasping_Wk11<-factor(data_clasping$Clasping_Wk11, ordered = TRUE , levels = c(0:4))
data_clasping$Clasping_Wk12<-factor(data_clasping$Clasping_Wk12, ordered = TRUE , levels = c(0:4))

colnames(data_clasping)[which(names(data_clasping) == "Gender")] <- "Sex"
Clasping_Week6<-summary(data_clasping$Clasping_Wk6)
Clasping_Week7<-summary(data_clasping$Clasping_Wk7)
Clasping_Week8<-summary(data_clasping$Clasping_Wk8)
Clasping_Week9<-summary(data_clasping$Clasping_Wk9)
Clasping_Week10<-summary(data_clasping$Clasping_Wk10)[-6]
Clasping_Week11<-summary(data_clasping$Clasping_Wk11)
Clasping_Week12<-summary(data_clasping$Clasping_Wk12)
count<-rbind(Clasping_Week6,Clasping_Week7,Clasping_Week8,Clasping_Week9,Clasping_Week10,Clasping_Week11,Clasping_Week12)
count
data_c<-data.frame(Week=c(6:12),count) %>% gather(Clasping, value, X0:X4)
data_c$Clasping<-factor(data_c$Clasping, levels = c("X0","X1","X2","X3","X4"), labels = c(0:4), ordered = TRUE)

```

```{r barplot_clas, echo = FALSE, fig.cap ='Bar plot for clasping over the 5 weeks', fig.pos="H",  fig.width = 10, fig.height = 9,  out.width = "1\\textwidth"}
ggplot(data_c, aes(fill=Clasping, y=value, x=Week)) + 
    geom_bar( stat="identity", position = position_stack(reverse = TRUE)) +
    scale_fill_viridis(discrete = T)+ylab("Count")
```
\newpage
- Cumulative Link Mixed Models are used to model clasping variable

We will fit the following cumulative link mixed model to the Clasping variable($Y_i$):

$logit(P(Y_i\leq j))=\theta_j-\beta_1(Genotype_i)-\beta_2(Housing_i)-\beta_3(Sex_i)-\beta_4(Week_i)-u(Box_i)$\\
where $i=1,...,973$ and $j=0,...,4$.


```{r echo=FALSE}
long_data_clasping<-data_clasping %>% 
  gather(v, value, Clasping_Wk6:Clasping_Wk12) %>% 
  separate(v, c("col", "Week"),sep="_Wk") %>% 
  arrange(MOUSE_ID) %>% 
  spread(col, value)


long_data_clasping$Week <-as.numeric(long_data_clasping$Week)
long_data_clasping_sorted<-long_data_clasping[order(long_data_clasping$MOUSE_ID,
                                                    long_data_clasping$Box,long_data_clasping$Week),]
long_data_clasping_sorted$Clasping<-factor(long_data_clasping_sorted$Clasping,ordered = TRUE , levels = c(0:4))
#library(ordinal)
mod = clmm(Clasping~Genotype + Housing+ Sex+Week+(1|Box)
           , data=long_data_clasping_sorted, Hess=TRUE)
summary(mod)

```

The coefficients of Genotype HD, Male and Week are positive indicating that the clasping rate being higher is more likely for higher weeks, among males and for HD group. The odds ratio of clasping rate being in category $j$ or above for male to female is 1.84. 

```{r barplot_clas_sex, echo = FALSE, fig.cap ='Bar plot for clasping over the 5 weeks between sex', fig.pos="H",  out.width = "0.6\\textwidth"}
data_c_group<-data.frame(long_data_clasping_sorted %>% filter(complete.cases(.)) %>%
   group_by(Week,Clasping, Sex, Genotype,Housing) %>% 
     tally())
ggplot(data_c_group, aes(fill=Clasping, y=n, x=Week)) + 
    geom_bar( stat="identity", position = position_stack(reverse = TRUE)) +
    scale_fill_viridis(discrete = T)+facet_wrap(~Sex)+ylab("Count")
```

```{r barplot_clas_geno, echo = FALSE, fig.cap ='Bar plot for clasping over the 5 weeks between genotype', fig.pos="H",  out.width = "0.6\\textwidth"}
ggplot(data_c_group, aes(fill=Clasping, y=n, x=Week)) + 
    geom_bar( stat="identity", position = position_stack(reverse = TRUE)) +
    scale_fill_viridis(discrete = T)+facet_wrap(~Genotype)+ylab("Count")
```
# Post-Hoc test for Clinical variables

Except for the clasping score,  we modelled all other clinical variables using Linear Mixed Models (LMMs). Since the clasping score was an ordinal response, a Cumulative Link Mixed Model was used instead of an LMM.

Example of interpretation for Table \ref{tab:brain_weight}: Among different housing types, SH showed the highest number of significant pairwise differences in brain weight among different sex and genotype contrasts. With the exception of housing type EX, both the other housing types showed a significant increase in average brain weight in WT female mice compared to HD counterparts. In addition, HD male mice have greater brain weight than HD female mice on average in SH housing.  

Figure 1 illustrates the pairwise comparisons. The blue bars are the confidence intervals for the estimated marginal means. However, pairwise comparisons are done using the red arrows, instead of using the confidence intervals directly. Pairwise comparison is considered to be significant if a red arrow from contrast does not overlap with another contrasts arrow. For instance in Figure \ref{fig:emmeans}, the two red arrows from WT male and HD female mice do not overlap in SH housing and are thus deemed significant.

```{r, echo=FALSE}
X1 <- lme(Brain_Weight ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Brain_Weight)))
X2 <- lme(DG_Swing_Fore ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Swing_Fore)))
X3 <- lme(DG_Swing_Hind ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Swing_Hind)))
X4 <- lme(DG_Stride_Length_Fore ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Stride_Length_Fore)))
X5 <- lme(DG_Stride_Length_Hind ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Stride_Length_Hind)))
X6 <- lme(DG_Stance_Width_Hind ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Stance_Width_Hind)))
X7 <- lme(DG_Propel_Brake_Ratio_FORE ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(DG_Propel_Brake_Ratio_FORE)))
X8 <- lme(Butyrate ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Butyrate)))
X9 <- lme(Valerate ~ Genotype + Housing+ Sex+
            Genotype:Sex+Sex:Housing+Housing:Genotype,
          random =(~1|Box), data = subset(data_clean, !is.na(Valerate)))
X10 <- lme(Water_Intake ~ Genotype + Housing+ Sex+Week+
            Genotype:Sex+Sex:Housing+
            Genotype:Week+ Sex:Week + Housing:Week+Housing:Genotype,
          random =(~1|Box), data = subset(food_sorted, !is.na(Water_Intake)))

X11 <- lme(Weight ~ Genotype + Housing+ Sex+Week+
            Genotype:Sex+Sex:Housing+
            Genotype:Week+ Sex:Week + Housing:Week+Housing:Genotype,
          random =(~1|Box), data = subset(long_data_sorted, !is.na(Weight)))
X12 <- lme(Rotarod ~ Genotype + Housing+ Sex+Week+
            Genotype:Sex+Sex:Housing+
            Genotype:Week+ Sex:Week + Housing:Week+Housing:Genotype,
          random =(~1|Box), data = subset(rotarod_data, !is.na(Rotarod)))


#stance widtg forepaw, propel:brake ratio forepaw, buterate, valerate
```

## Linear Mixed Model
### Brain weight

```{r brain_weight, echo=FALSE}
x<-emmeans(X1, pairwise ~ Genotype*Sex|Housing)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex",caption = "Post-hoc results for genotype and sex given housing type in variable brain weight") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(7,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

```{r emmeans, echo = FALSE, fig.cap ='Graphical comparisons from emmeans method', fig.pos="H",  out.width = "0.9\\textwidth"}
plot(x, comparisons = TRUE)
```

### Swing time forepaw
```{r, echo=FALSE}
x<-emmeans(X2, pairwise ~ Genotype|Housing*Sex)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(8,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

### Swing time hindpaw
```{r, echo=FALSE}
x<-emmeans(X3, pairwise ~ Genotype|Housing*Sex)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(8,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

### Stride length forepaw
```{r, echo=FALSE}
x<-emmeans(X4, pairwise ~ Genotype|Housing*Sex)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(8,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

### Stride length hindpaw
```{r, echo=FALSE}
x<-emmeans(X5, pairwise ~ Housing|Genotype*Sex)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(8,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

```{r, echo=FALSE}
x<-emmeans(X5, pairwise ~ Housing|Genotype)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(7,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

### Stance width forepaw
```{r, echo=FALSE}
x<-emmeans(X6, pairwise ~ Genotype|Housing*Sex)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(8,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

### Propel:brake ratio forepaw
```{r, echo=FALSE}
x<-emmeans(X7, pairwise ~ Genotype|Housing*Sex)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(8,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

### Butyrate
```{r, echo=FALSE}
x<-emmeans(X8, pairwise ~ Sex|Housing)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
  column_spec(7,bold = that_cell,
              color =  ifelse(that_cell,"red","black"))%>%
  column_spec(1,color =  ifelse(that_cell,"red","black"))
```

### Valerate

```{r, echo=FALSE}
x<-emmeans(X9, pairwise ~ Sex|Housing)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
  column_spec(7,bold = that_cell,
              color =  ifelse(that_cell,"red","black"))%>%
  column_spec(1,color =  ifelse(that_cell,"red","black"))
```

### Water Intake
```{r, echo=FALSE}
x<-emmeans(X10, pairwise ~ Genotype*Housing*Sex)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(6,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

### Weight

```{r, echo=FALSE}
x<-emmeans(X11, pairwise ~ Genotype*Housing*Sex)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(6,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

### Rotarod

```{r, echo=FALSE}
x<-emmeans(X12, pairwise ~ Genotype*Housing|Sex)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(7,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

## Cumulative Link Mixed Model
### Clasping Score

```{r, echo=FALSE}
x<-emmeans(mod, pairwise ~ Genotype*Sex|Housing)
df<-data.frame(x$contrasts)
that_cell<-df$p.value<0.05
kbl(x$contrasts, longtable=T, linesep = "", booktabs = T, digits = c(4,4,4,4,4,4),format = "latex") %>%
  kable_styling(latex_options = c("repeat_header"))%>%
column_spec(7,bold = that_cell,
            color =  ifelse(that_cell,"red","black"))%>%
column_spec(1,color =  ifelse(that_cell,"red","black"))
```

# References {-}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\noindent
<div id="refs"></div>
```{r refmgr references, results="asis", echo=FALSE}
# Print
```
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\setlength{\parskip}{0pt}

# R session information {-}

```{r}
sessionInfo()
```